Prey skill shapes the acquisition of predator expertise in a virtual bi-trophic system

### 2025-05-25 manuscript/manuscript.Rmd
Linux --> .rmd manuscript update up to results. need to finish results

---
title: "Prey movement shapes the acquisition of predator expertise in a virtual bi-trophic system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

@@

dat <- fread(
  file.path(
    path, "FraserFrancoetal2023-data.csv"
  ),
  select = c(
    "timestamp", "match_id", "realm_id",
    "environment_id", "avatar_id", "predator_id",
    "game_duration", "total_xp_pred", "cumul_xp_pred",
    "hunting_success", "pred_speed", "prey_avg_speed"
  )
)

@@

The acquisition of expertise is crucial for predators to be successful hunters. To achieve this, predators must hone their skills and gain knowledge through repeated and extensive practice. On the other hand, prey may hinder the acquisition of predator expertise by employing antipredator tactics to evade detection and pursuit. However, empirical evidence on how predators acquire expertise through repeated encounters with their prey remains limited, largely due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in human predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience, indicating that they acquired expertise through extensive practice. At the population-level, we found that faster prey impaired the acquisition of expertise by reducing hunting success. Prey speed was also an important mediator of this relationship at the individual level, driving differences among predators in the acquisition of expertise. Our study outlines how prey antipredator behaviour can mediate the acquisition of expertise in predator populations.

@@

Predation is a fundamental biological process that acts as an agent of evolutionary change and shapes community structure [@abrams2000;@wirsing.etal2021]. By consuming prey, predators regulate prey populations, limit the spread of diseases, and mediate energy flow across trophic levels [@Hairston.1993;@ripple.etal2014;@schmitz2017]. These processes emerge from interactions between individual predators and their prey, lying at the core of key concepts such as predation risk and prey selection [@pettorelli.etal2015;@labarge.etal2024]. Predators also respond to interactions with their prey, increasing their effectiveness by adapting their strategy and choosing suitable habitats to hunt [@quevedo.etal2009;@schmitz2017]. Central theory suggests that these responses by predators may result from acquiring expertise through practicing and learning the proper skills to locate, select, and capture their prey [@woo.etal2008;@wooster.etal2023;@labarge.etal2024]. Hence, the role of expertise remains a missing but essential link to assess how individual characteristics shape prey consumption during predator-prey interactions.

Expertise can be defined as the characteristics, skills, and knowledge that provide individuals with the ability to outperform novices on complex tasks [@dukas2017]. Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by associative learning, by developing search images, or by exploiting cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Through these processes, expert predators should have greater knowledge, better energy management, and acute motor skills that increase their chances of locating and capturing prey [@dukas2019]. Therefore, differences among predators in their capacity to acquire expertise throughout their lifetime may underlie differences in hunting success. This is an important topic of study since predator cognition can maintain behavioural variation, shape prey phenotypic composition, and destabilize predator-prey systems [@kondoh2010;@skelhornandcandy2016;@kikuchi.etal2023].

Prey use antipredator tactics such as camouflage to avoid detection and rapid escapes to evade capture [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. These strategies are hypothesized to drive differences in prey consumption among predators by disrupting the acquisition of hunting expertise [@wooster.etal2023]. Experimental studies have shown that certain camouflage tactics of prey can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013]. For example, @troscianko.etal2018 found that disruptive colouration interfered with human subjects' ability to form search images, hindering improvements in detection times over repeated attempts. Thus, antipredator behaviour may also hold the potential to hinder the acquisition of predator expertise. For instance, some predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which could impair their acquisition of hunting expertise. To our knowledge, there is no empirical evidence showing links between prey antipredator behaviour and the acquisition of expertise in human and nonhuman predators. This represents a significant gap in our understanding of predator-prey interactions, as individual variation in predator expertise could lead to shifts in foraging behaviour based on whether predators select prey for their frequency or their traits [@allen.etal1988;@ishii.shimada2012;@wooster.etal2023].

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Here we mitigate these challenges by using *Dead by Daylight* as our study system, a videogame where four prey players must forage for resources while avoiding predation by a fifth player. Similar to agent-based simulations, *Dead by Daylight* provides controlled virtual environments to test ecological hypotheses [see @lymbery.etal2023 for an example with *Age of Empires II*], but with the advantage of having real players that interact in the virtual space. In this game, the predator population comprises individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also elicits natural reactions in players such as freezing when predation is imminent (M.F.F., personal observations), corroborating another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological phenomena can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on thousands of interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, *Dead by Daylight* allows us to tackle fundamental questions about the role of antipredator behaviour and experience in predator-prey interactions.

In this study, we assess how repeated encounters with prey shapes predator hunting success using data from players in *Dead by Daylight*. We quantify expertise acquisition as the relationship between hunting success (i.e. the probability of capturing all prey) and repeated experience (i.e. cumulated atempts). First, we test the hypothesis that predator success will increase with experience up to some point at which it will stabilize [@dukas2019]. However, we expect this pattern to change depending on the movement of the prey encountered. We hypothesize that prey will influence expertise acquisition, and predict that faster prey will reduce the gain in expertise. Therefore, we also investigate how prey movement influences expertise acquisition at the individual level. If prey speed does not influence hunting success, we predict that the gain in expertise will be similar among individuals. Alternatively, if prey speed influences hunting success, then the acquisition of expertise will vary among individuals.

@@

Behaviour Interactive Inc. provided data that spanned six months of gameplay recorded for every player from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We analyzed only matches where players did not know each other. We filtered any matches where players were inactive, such as when mean distances travelled per second (i.e. speed) were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as intended.

We sampled players that played 300 matches or more and monitored all their matches from the first to a maximum of 500 matches. We recognize that we could have introduced a bias by retaining only those individuals, as they might already be seasoned video game enthusiasts and exhibit expert-level performance in their early matches in *Dead by Daylight*. Thus, we verified that our sample was not biased by comparing a random sample of players that played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe as our sampled population. We then took the first 20 matches played by these players, including those from our sampled population, and compared their median hunting success using a Bayesian hierarchical linear model. We found that all four groups had similar success as predators (Appendix 1: Table S1 and Figure S1), suggesting an absence of bias due to data sampling.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` players who played as the predator, with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience, the mean speed of the groups of prey that the predator player encountered, and the mean rank of the prey encountered (a proxy for prey skill). The ranking system in *Dead by Daylight* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank), and failing to account for it would prevent us from detecting a change in the predator's foraging success with experience.

We analyzed the mean speed of the prey group encountered by the predator. We measured the prey's speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We did not account for matches where predators played as the prey.

@@

We tested how predators developed their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @pedersen.etal2019. The first model (I) was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same acquisition of expertise, with the model estimating a trend for the average individual (i.e. global smoother). The second model (II) included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model (III), we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth (IV) and fifth (V) models were expansions of the third and second models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

We computed the five models using a modified version of the beta-binomial distribution. Hunting success was estimated as the probability of consuming the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link. We used ten basis functions (K = 10) for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We fitted all models in R (version 4.1.2) using Hamiltonian Monte Carlo (HMC) sampling with the package "brms" version 2.17.0 [@burkner2017], an R front-end for the STAN software [@StanDevelopmentTeam2023], and "cmdstanr" version 0.4.0 [@gabry.cesnovar2021] as the back-end for parameter estimation (cmdstan installation version 2.28.2). For further details, please consult the GitHub repository of this project (https://github.com/quantitative-ecologist/predator-expertise).

We used weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @fraserfranco.etal2022, we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$), a positive Gaussian prior ($N(1,\ 0.5^{2}))$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5^{2})$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$). We compared the predictive accuracy of all five models using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@vehtari.etal2017;@piironen.vehtari2017;@vehtari.etal2022].

@@

We parametrized the GAMMs to run four chains. We ran 2500 iterations with a thinning set to eight for model I, and 1500 iterations with a thinning set to four for the remaining additive models (models II to V). We set the first 500 iterations of each model as warm-ups. For each model, we obtained 1000 posterior samples per parameter. We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models.

@@

We tested the hypothesis that antipredator behaviour impairs the acquisition of expertise at the population level by visually comparing the global trends of model II (not controlling for prey speed) and model V (controlling for prey speed) relating hunting success to cumulated experience (Figure 1). At the individual level, we tested our hypothesis that antipredator behaviour generates differences among predators in expertise acquisition by comparing individual-level parameters of model II and model V. Specifically, we compared the standard deviations of 1) the random intercepts (i.e. mean differences in hunting success), 2) the random slopes (i.e. linear component relating hunting success with experience), and 3) the curve wiggliness (i.e. nonlinear component relating hunting success with experience).

@@

Out of all five GAMM models, the two that accounted for the prey group's rank and speed were the best at predicting the data with similar predictive accuracies (Table 1). Models in which prey rank was not accounted for resulted in no change in hunting success with experience for the average individual (i.e. no gain in expertise, results not shown). Accounting for the prey rank resulted in a concave-shaped relationship (model II), with the highest success ranging between ~200 and ~300 matches (Figure 1A).

We found strong evidence of a negative relationship between hunting success and prey speed (Figure S2). As predicted, the effect of experience on hunting success for the average individual followed a diminishing returns curve when controlling for prey speed (model V), with predators optimizing their success after playing ~300 matches (Figure 1C). The curve shows a 38% increase in the probability of consuming all prey for the average individual between the first and the ~330$^{th}$ match, where success reached a plateau (Figure 1C).

@@

Prey speed did not influence among-individual differences in average hunting success as the posterior distributions of the standard deviations of individual intercepts were almost completely overlapping (Figure 2, median = `r toString(round(sum1[2,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[2,]$Estimate, digits = 2))`). However, individuals differed in the acquisition of their hunting expertise (Figure 1B-D). We found strong evidence that the speed of the prey mediated among-individual differences in the linear relationship between success and experience, as there were substantial differences in the standard deviations of the individual slopes between the two models (Figure 2, median = `r toString(round(sum1[3,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[3,]$Estimate, digits = 2))`). Differences among individuals in the direction of the linear relationship between success and experience were `r toString(round(sum1[3,]$Estimate / sum2[3,]$Estimate, digits = 2))` times lower when we removed the effect of prey speed (i.e. accounting for it in the model).

@@

@@

```{r fig2, out.width = "100%", fig.cap = "**Figure 2**. Posterior distributions of the standard deviation of individual-level parameters estimated by the GAMM. The parameters are displayed on the y axis, and their standard deviation are displayed on the x axis. The intercept and slope standard deviations refer to the linear components of the estimated relationship between hunting success and cumulative experience. The standard deviation of the wiggliness parameter refers to the shape of the curves (i.e. nonlinear component). The vertical lines are the medians of the posterior distributions. For each panel, the light-coloured distributions (i.e. green) are for the model with a shared trend where we did not account for prey speed (model II), and the darker-coloured distributions (i.e. purple) are for the model with a shared trend where we accounted for prey speed (model IV)."}

>>>>

---
title: "Prey skill shapes the acquisition of predator expertise in a virtual bi-trophic system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

@@

dat <- fread(
  file.path(
    path, "FraserFrancoetal2023-data.csv"
  ),
  select = c(
    "timestamp", "match_id", "realm_id",
    "environment_id", "avatar_id", "predator_id",
    "game_duration", "total_xp_pred", "cumul_xp_pred",
    "hunting_success", "pred_speed", "prey_avg_speed", 
    "prey_avg_rank"
  )
)

@@

The acquisition of expertise is crucial for predators to be successful hunters. To achieve this, predators must hone their skills and gain knowledge through repeated and extensive practice. Prey can interfere with this process and hinder the acquisition of predator expertise by employing antipredator tactics to evade detection and pursuit. However, empirical evidence on how predators acquire expertise through repeated encounters with their prey remains limited, largely due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in human predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience, indicating that they acquired expertise through extensive practice. At the population-level, we found that more skillful prey impaired the acquisition of expertise by reducing hunting success. Prey skill was also an important mediator of this relationship at the individual level, driving differences among predators in the acquisition of expertise. More skillful prey moved faster, and predators were less successful against faster prey. Our results suggest that skilled prey use behaviours to outmanoeuver their predators, thereby mediating how predators acquire expertise at the population and individual level.

@@

Predation is a fundamental biological process acting as an agent of evolutionary change that regulates prey populations, limits the spread of diseases, and mediates energy flow across trophic levels [@hairston1993;@abrams2000;@ripple.etal2014;@wirsing.etal2021]. These processes are driven by changes in predation risk and prey selection across landscapes as predators adjust their strategy and select suitable habitats to hunt [@quevedo.etal2009;@pettorelli.etal2015;@labarge.etal2024;@schmitz2017]. Central theory suggests that predators become more efficient hunters by acquiring expertise through practicing and learning the proper skills to locate, select, and capture their prey [@woo.etal2008;@wooster.etal2023;@labarge.etal2024]. Therefore, differences in hunting success between predators may be attributed to differences in the acquisition of hunting expertise. Unraveling how this expertise emerges remains a missing, but essential, link to assess how foraging behaviour shapes prey consumption during predator-prey interactions.

Expertise can be defined as the characteristics, skills, and knowledge that provide individuals with the ability to outperform novices on complex tasks [@dukas2017]. Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by associative learning, by developing search images, or by exploiting cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Through these processes, expert predators should have greater knowledge, better energy management, and acute motor skills that increase their chances of locating and capturing prey [@dukas2019]. Theory predicts that predator cognition can maintain behavioural variation, shape prey phenotypic composition, and destabilize predator-prey systems [@kondoh2010;@skelhornandcandy2016;@kikuchi.etal2023]. Therefore, individual differences among predators in their capacity to acquire expertise over time may be important for the ecological consequences of predation.

Prey use antipredator tactics such as camouflage to avoid detection and rapid escapes to evade capture [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. These strategies are hypothesized to drive differences in prey consumption among predators by disrupting the acquisition of hunting expertise [@wooster.etal2023]. Experimental studies have shown that certain camouflage tactics of prey can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013]. For example, @troscianko.etal2018 found that disruptive colouration interfered with human subjects' ability to form search images, hindering improvements in detection times over repeated attempts. Antipredator behaviour may also hold the potential to hinder the acquisition of predator expertise. For instance, some predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which could impair their acquisition of hunting expertise if they chase faster prey. Moreover, prey antipredator behaviour can improve with experience through continued exposure to predators [@turner2006;@kelley.magurran2011;@lonnstedt.etal2012]. Indeed, animal reintroduction programs show that individuals trained to recognize predators display more antipredator behaviours and are 1.5 times more likely to survive than predator-naive individuals [@tetzalf.etal2019]. To our knowledge, there is little empirical evidence showing links between prey skill, antipredator behaviour, and the acquisition of predator expertise in human and nonhuman predators, representing a significant gap in our understanding of predator-prey interactions.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Here we mitigate these challenges by using *Dead by Daylight* as our study system, a videogame where four prey players must forage for resources while avoiding predation by a fifth player. Similar to agent-based simulations, *Dead by Daylight* provides controlled virtual environments to test ecological hypotheses [see @montiglio.etal2025 for a review], with the advantage of having real players that interact in the virtual space. In this game, the predator population comprises individuals that either ambush or hunt at high speeds (i.e., mean movement speed along an ambush-cursorial continuum of tactics), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also elicits natural reactions in players such as freezing when predation is imminent (M.F.F., personal observations), corroborating virtual ecological studies showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological phenomena can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on thousands of interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, *Dead by Daylight* allows us to tackle fundamental questions about the role of antipredator behaviour and experience in predator-prey interactions.

In this study, we assess how repeated encounters with prey shapes predator hunting success using data from players in *Dead by Daylight*. We quantify expertise acquisition as the relationship between hunting success (i.e. the probability of capturing all prey) and repeated experience (i.e. cumulated matches). First, we test the hypothesis that predator success will increase with experience up to some point at which it will stabilize [@dukas2019]. However, we hypothesize that prey skill and speed will influence predator expertise acquisition, and predict that faster and more skillful prey will reduce the gain in expertise at the population level. Prey with greater skills should move at faster speeds. Moreover, if prey skill and speed influence hunting success at the population level, then the acquisition of predator expertise should vary among individuals given these two prey features. Alternatively, if prey characteristics do not influence predator hunting success at the population level, we predict that the gain in predator expertise will be similar among individuals.

@@

Behaviour Interactive Inc. provided data that spanned six months of gameplay recorded for every player from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))` (game build 4.4.0 to 4.7.2). We analyzed only matches where players did not know each other and were unable to communicate using voice-recording devices. We filtered any matches where players were inactive, such as when mean distances travelled per second (i.e. speed) were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as intended.

Our study population comprised `r toString(format(length(unique(dat$predator_id))))` players that played at least 300 matches. We monitored all their matches from the first to a maximum of 500 matches, with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches overall. The predator-players' experience in our population sample varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). We recognize that we could have introduced a bias by retaining only those individuals, as they might already be seasoned video game enthusiasts and exhibit expert-level performance in their early matches in *Dead by Daylight*. Thus, we tested sample bias by comparing a random sample of players that played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe as our population sample. We then took the first 20 matches played by these players, including those from our population sample, and compared their mean hunting success using a Bayesian hierarchical linear model. We found that all four groups had similar success as predators in their first 20 matches (Appendix 1: Table S1 and Figure S1), suggesting an absence of bias due to data sampling.

We collected the following information for every match in our population sample : the player's anonymous ID, the predator player's hunting success, the predator player's cumulated experience, and the mean rank and speed of the the group of prey that the predator player encountered. We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). We defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. Laslty, we measured the prey's rank as the mean rank of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_rank, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_rank, na.rm = T), digits = 2), nsmall = 2))`), and the prey's speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s).

We used the mean rank of the prey as a proxy of their skill. The ranking system in *Dead by Daylight* varies between `r toString(format(min(dat$prey_avg_rank)))` and `r toString(format(max(dat$prey_avg_rank)))` (1 indicating the highest skill), and was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). The skill of a player increases based on their performance from match to match. While the ranking system represents an approximation of prey skill, it was the most readily available metric we could access to determine the skill of the prey group. In addition, the pairing system is subject to variation depending on factors such as player availability, which can result in predator-prey groups with unbalanced skill. This allowed us to evaluate the effect of the prey's skill on predator success.

@@

We tested how predators developed their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success and the predators' cumulative experience. We parametrized the models following the method of @pedersen.etal2019. For the first and second models (I and II), we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. Both models included a common global smoothing function (i.e., population effects) and random smooths for the predator ID (i.e., individual effects). Model I included the standardized match duration as a covariate, while model II included the standardized match duration and the average rank of the prey group, allowing us to test if the variation among predators in expertise acquisition was driven by prey skill. In the third model (III), we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. In this model, we control for the standardized match duration and the average rank of the prey group. The fourth (IV) and fifth (V) models were expansions of the model III and model II respectively, where we included the standardized prey speed to assess its effect on the relationship between success and experience.

We computed the five models using a modified beta-binomial distribution. Hunting success was estimated as the number of prey consumed out of four, with probability of success drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor. The precision parameter ($\phi$) was estimated using a log-link function. We used ten basis functions (K = 10) for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We compared the predictive accuracy of all five models using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@vehtari.etal2017;@piironen.vehtari2017;@vehtari.etal2022].

We defined weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @fraserfranco.etal2022, we defined a positive Gaussian prior ($N(1,\ 0.5^{2}))$) for the slope of the game duration because longer trials lead to greater success, a Gaussian prior ($N(0,\ 1^{2}))$) for the slope of the prey rank, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5^{2})$) because encountering faster prey is associated with lower success in this system [@fraserfranco.etal2022]. We defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$). We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$).

To test the hypothesis that prey speed increases with prey skill, we fitted a Bayesian linear regression estimating the relationship between prey speed and the standardized prey rank. We controlled for the game duration, using it as a linear covariate in the model, and included the predator identity as a random effect. We used a weakly informative Gaussian prior on the slope of the game duration ($N(0,\ 0.5^{2})$) and a weakly informative negative Gaussian prior the slope of the prey rank ($N(-0.5,\ 1^{2})$) because prey of greater skill (i.e., rank 1 is the highest skill) should move faster. Lastly, we used a weakly informative Gaussian prior on the intercept ($N(2,\ 0.5^{2})$) because prey should move at around two meters per second, and half-Gaussian priors on the standard deviation parameters ($N(0,\ 0.5^{2})$).

@@

We parametrized the GAMMs to run four chains of 1500 iterations each, sampling one parameter value every four iterations. We set the first 500 iterations of each model as warm-ups. Similarly, the linear model ran four chains of 1500 iterations each, sampling a parameter value at every iteration, and discarding the first 500 iterations. For GAMM model, we obtained 1000 posterior samples per parameter, and 4000 values for the linear model. We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models.

We fitted all models in R (version 4.1.2) using Hamiltonian Monte Carlo (HMC) sampling with the package "brms" version 2.22.0 [@burkner2017], an R front-end for the STAN software [@StanDevelopmentTeam2023], and "cmdstanr" version 0.8.0 [@gabry.cesnovar2021] as the back-end for parameter estimation (cmdstan installation version 2.35.0). For further details on how to reproduce the results, please consult the GitHub repository of this project (https://github.com/quantitative-ecologist/predator-expertise).

@@

We tested the hypothesis that prey skill and antipredator behaviour impairs the acquisition of expertise at the population level by visually comparing the global trends of model I (only controlling for game duration), model II (controlling for game duration and prey rank) and model V (controlling game duration, prey rank, and prey speed) (Figure 1). At the individual level, we tested our hypothesis that prey skill and antipredator behaviour generate differences among predators in expertise acquisition by comparing the individual-level variance parameters. Specifically, we compared the standard deviations of 1) the random intercepts (i.e. mean differences in hunting success), 2) the random slopes (i.e. linear component relating hunting success with experience), and 3) the curve wiggliness (i.e. nonlinear component relating hunting success with experience).

@@

Out of all five GAMM models, those that included a global smoother were better at predicting the data compared to those without a global smoother, indicating that predators shared a similar relationship between prey capture and experience (Table 1). Model V, which accounted for the prey group's rank and speed was the best at predicting the data (Table 1).

@@

Models in which prey rank was not accounted for resulted in no change in hunting success with experience for the average individual (i.e. no gain in expertise, Figure 1A). As we predicted, accounting for the prey rank resulted in a diminishing returns curve (model II), with predators optimizing their success after playing ~400 matches (Figure 1B). The predictions indicate a 40% increase in the probability of consuming all prey for the average individual between the first and the ~400$^{th}$ match, where success reached a plateau (Figure 1B). Controling for prey speed did not influence the results at the population level (Figure 1C). Yet, we found strong evidence of a negative relationship between hunting success and prey speed, indicating that hunting faster prey reduced the predator's chances of capture (Figure S2). Moreover, prey with more expertise moved at higher speeds compared to less skilled prey (Figure 2).

@@

Prey speed did not influence among-individual differences in average hunting success as the posterior distributions of the standard deviations of individual intercepts were almost completely overlapping (Figure 3, median = `r toString(round(sum1[2,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[2,]$Estimate, digits = 2))`). However, individuals differed in the acquisition of their hunting expertise (Figure 1B-D). We found strong evidence that the speed of the prey mediated among-individual differences in the linear relationship between success and experience, as there were substantial differences in the standard deviations of the individual slopes between the two models (Figure 3, median = `r toString(round(sum1[3,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[3,]$Estimate, digits = 2))`). Differences among individuals in the direction of the linear relationship between success and experience were `r toString(round(sum1[3,]$Estimate / sum2[3,]$Estimate, digits = 2))` times lower when we removed the effect of prey speed (i.e. accounting for it in the model).

```{r fig2, out.width = "100%", fig.cap = "**Figure 2**. Median posterior predictions of the relationship between prey speed and prey skill. The average speed of the prey group is on the y-axis, and the average rank of the prey group is on the x-axis. The colour gradient showcases the density of observations, with lighter colours indicating lower densities, and darker colours indicating higher densities of observations. Lower rank values indicate that the prey group is more skilled, while higher values indicate that the prey are less skilled"}

knitr::include_graphics(
  here::here("outputs", "outputs_figures", "figurex.png")
)
```

@@

```{r fig3, out.width = "100%", fig.cap = "**Figure 3**. Posterior distributions of the standard deviation of individual-level parameters estimated by the GAMM. The parameters are displayed on the y axis, and their standard deviation are displayed on the x axis. The intercept and slope standard deviations refer to the linear components of the estimated relationship between hunting success and cumulative experience. The standard deviation of the wiggliness parameter refers to the shape of the curves (i.e. nonlinear component). The vertical lines are the medians of the posterior distributions. For each panel, the light-coloured distributions (i.e. green) are for the model with a shared trend where we did not account for prey speed (model II), and the darker-coloured distributions (i.e. purple) are for the model with a shared trend where we accounted for prey speed (model IV)."}
